---
title: "workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(gmhdatahub)
```

## Understanding Key Terms in API Data Processing

Processing API response data involves a series of steps and concepts that ensure the data is correctly received, interpreted, transformed, and utilized within an application.

Below is an overview of key terms and their roles within the typical workflow of handling API responses:

### API Response

The data returned by an API after a request is made. It usually comes in formats like JSON, XML, or others.

### Parsing

- Definition: Analyzing the API response data structure to convert it into a usable format within your application.

- Example: Converting JSON response data into a list or data frame in R.

### Serialization & Deserialization

- **Serialization:** Converting an in-memory data structure into a format that can be easily stored or transmitted (e.g., JSON, XML).

- **Example:** Using `jsonlite::toJSON()` in R to convert an object into a JSON string before sending it in an API request.

- **Deserialization:** The inverse process; converting serialized data back into an in-memory data structure.

- **Example:** Parsing a JSON response using `jsonlite::fromJSON()` in R to convert it into a list or data frame.

### Marshaling & Unmarshaling

- **Marshaling:** Transforming memory objects to a format suitable for storage or transmission, often used in remote procedure calls (RPC).

- **Example:** Using Protocol Buffers to marshal data before sending it to a microservice.

- **Unmarshaling:** The reverse process; converting marshaled data back into memory objects.

- **Example:** Decoding Protocol Buffers data back into native objects in a service.

### Schemas

- **Definition:** Formal definitions of the structure, types, and constraints of data (e.g., JSON Schema, XML Schema).

- **Example:** Using a JSON Schema to validate that an API response contains required fields like `id`, `name`, and `email`.

### Encoding & Decoding

- **Encoding:** Converting data into a specific format for transmission or storage, ensuring compatibility or efficiency.

- **Example:** Encoding data as Base64 when embedding binary data in a JSON response.

- **Decoding:** The reverse process; converting encoded data back into its original format.

- **Example:** Decoding a Base64 string back into binary data after receiving it from an API.

### Encryption & Decryption

**Encryption:** Converting data into a secure format to prevent unauthorized access.

**Example:** Using HTTPS to encrypt data in transit or encrypting specific fields within an API response.

**Decryption:** Converting encrypted data back into its original, readable format.

**Example:** Decrypting a sensitive field in an API response using a predefined key.

### Extraction

Retrieving specific pieces of data from a larger dataset.

**Example:** Extracting the `user` object from a nested JSON API response.

Transformation
Modifying data from one format or structure to another to meet the application's requirements.

Example: Converting date strings from the API into JavaScript Date objects.

10. Enrichment
Enhancing the API response data by adding additional information from other sources.

Example: Adding geographic coordinates to user data by cross-referencing with a location service.

11. Validation
Checking that the API response data meets certain criteria or standards.

Example: Verifying that a numerical field falls within an expected range or that required fields are present.

12. Normalization
Structuring data to reduce redundancy and improve integrity.

Example: Splitting a full address into separate street, city, state, and zip fields.

13. Aggregation
Combining multiple data points into a summarized or consolidated form.

Example: Calculating the total sales from a list of sales transactions received from an API.

14. Caching
Storing API response data temporarily to improve performance and reduce redundant requests.

Example: Using an in-memory cache like Redis to store user profiles fetched from an API.

15. Error Handling
Managing and responding to errors that occur during the processing of API responses.

Example: Retrying a failed API request or logging an error for later analysis when parsing fails.

16. Logging and Monitoring
Recording and tracking the processing of API responses for debugging, auditing, and performance analysis.

Example: Logging the time taken to parse and process each API response or monitoring for a high rate of failed validations.

